Briefing paper on Ethnicity and COVID-19 status in Scotland

### 2020-05-14 Covid_ethnicity_Scotland.Rmd
updated for ethnicity report

output:
  pdf_document

@@

A previous comparison of ONOMAP against SMR01 records in Scotland concluded that SMR01 records with a missing ethnic group code were broadly representative of records with ethnicity recorded.  (Clark & Buchanan ISD 2017)
The classification given below follows the Census categories, collapsed into five groups: White, Black, South Asian, Chinese, Other.

# Results

## Agreement between ethnicity assigned by name classification (ONOMAP) and self-reported ethnicity in  SMR records

@@

**Amanda see line below for example of how to embed numbers from R in text**

Table \ref{tab:crosstab} shows that the ONOMAP name classification assigns to the same as self-reported category `r testpositives.ethnic[1, 1]`

of those who self-report their ethnicity as White, 92% of those who self-report as South Asian, 71% of those who self-report as Chinese, but only 21% of those who self-report as Black.

@@

@@

pander(testpositives.ethnic.smr, missing=".",
       split.tables=Inf,
       split.cells=rep(25, 3), 
       caption="Severity of disease in test-positive cases, by ethnicity assigned in SMR records")

@@

rownames(table.demog.aug) <- gsub("_", " ", rownames(table.demog.aug))
knitr::kable(table.demog.aug,
             escape=FALSE, 
             booktabs=TRUE,
             label="onomap",
             digits=c(0, 0, 0, 200, 0, 200), 
             col.names=c(clean.header(colnames(table.demog.aug)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of severe disease with demographic variables including ethnicity based on name classification") %>%
    column_spec(column=1, width="4.5cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    landscape()
```

rownames(table.ethnicsmr.aug) <- gsub("_", " ", rownames(table.ethnicsmr.aug))
knitr::kable(table.ethnicsmr.aug,
             escape=FALSE, 
             booktabs=TRUE,
             label="ethnicsmr",
             digits=c(0, 0, 0, 200, 0, 200), 
             col.names=c(clean.header(colnames(table.ethnicsmr.aug)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of severe disease with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
    column_spec(column=1, width="4.5cm")
```

As shown above the numbers of persons testing positive from non-White ethnic groups is low. There is no evidence that the frequency of severe disease among those who test positive is any higher in non-White groups than in Whites. The frequency of severe disease was slightly lower in South Asans but with these small numbers this estimate is unstable.

# Conclusion
Although the numbers are small, there is no evidence that the main ethnic minorities in Scotland are at higher risk of severe COVID-19 than the  White  population. The point estimates for severe disease are lower in the main minority ethnic groups than in Whites.  However power to detect anything other than very large differences is limited as yet given the low numbers of non-White persons tested positive.  Therefore we will continue to monitor the situation.

Recording of ethnicity in hospital admission does not provide complete coverage of the population at risk of COVID-19.  The coding is also incomplete with substantial missingness.  Merging ethnicity data from the Scottish Morbidity Record with name based classification would introduce differential misclassification of ethnicity by health status.  ONOMAP/ name based classification does not reliably distinguish people of Black-Caribbean heritage from people of European ancestry.  Access to primary care data would likely increase the completeness with which ethnicity is assigned but governance issues need resolution.

>>>>

output:
  pdf_document:
    keep_tex: true

@@

A previous comparison of ONOMAP against SMR01 records in Scotland concluded that SMR01 records with a missing ethnic group code were broadly representative of records with ethnicity recorded (Clark & Buchanan ISD 2017).  The classification given below follows the Census categories, collapsed into five groups: White, Black, South Asian, Chinese, Other.

# Results
## Agreement between ethnicity assigned by name classification (ONOMAP) and self-reported ethnicity in  SMR records

@@

Table \ref{tab:crosstab} shows that the ONOMAP name classification assigns to the same as self-reported category `r table.ethnic[1, 1]` of those who self-report their ethnicity as White, `r table.ethnic[5, 5]` of those who self-report as South Asian, `r table.ethnic[3, 3]` of those who self-report as Chinese, but only `r table.ethnic[2, 2]` of those who self-report as Black.

@@

rownames(table.severe.demog) <- gsub("_", " ", rownames(table.severe.demog))
knitr::kable(table.severe.demog,
             escape=FALSE, 
             booktabs=TRUE,
             label="onomap",
             col.names=c(clean.header(colnames(table.severe.demog)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of severe disease with demographic variables including ethnicity based on name classification") %>%
    column_spec(column=1, width="4.5cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    landscape()
```

rownames(table.ethnicsmr.aug) <- gsub("_", " ", rownames(table.ethnicsmr.aug))
knitr::kable(table.ethnicsmr.aug,
             escape=FALSE, 
             booktabs=TRUE,
             label="ethnicsmr",
             col.names=c(clean.header(colnames(table.ethnicsmr.aug)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of severe disease with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
    column_spec(column=1, width="4.5cm")
```

As shown above the numbers of persons testing positive from non-White ethnic groups is low. There is no evidence that the frequency of severe disease among those who test positive is any higher in non-White groups than in Whites. The frequency of severe disease was lower in South Asans but with these small numbers this estimate is unstable.

## Relation of hospitalised COVID-19 to ethnicity in the Scottish population

```{r onomap.hosp, echo=FALSE, warning=FALSE, message=FALSE}

rownames(table.hospitalized.demog) <- gsub("_", " ", rownames(table.hospitalized.demog))
knitr::kable(table.demog.aug,
             escape=FALSE, 
             booktabs=TRUE,
             label="onomap.hosp",
             col.names=c(clean.header(colnames(table.hospitalized.demog)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of hospitalized disease with demographic variables including ethnicity based on name classification") %>%
    column_spec(column=1, width="4.5cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    landscape()
```

```{r ethnicsmr.hosp, echo=FALSE, warning=FALSE, message=FALSE}

rownames(table.hospitalized.demog.ethnicsmr) <- gsub("_", " ", rownames(table.hospitalized.demog.ethnicsmr))
knitr::kable(table.hospitalized.demog.ethnicsmr,
             escape=FALSE, 
             booktabs=TRUE,
             label="ethnicsmr.hosp",
             col.names=c(clean.header(colnames(table.hospitalized.demog.ethnicsmr)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of hospitalized disease with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
    column_spec(column=1, width="4.5cm")
```

# Conclusion
Although the numbers are small, it is possible to establish that South Asians in Scotland are at lower risk of severe disease than the White population.

For the Black population, reliable classification of ethnicity is restricted to those who have an SMR record, and the numbers of cases and controls in this group are too small for rate ratios to be reliably estimated.

ONOMAP/ name based classification does not reliably distinguish people of Black-Caribbean heritage from people of European ancestry.  Recording of ethnicity in hospital admission distinguishes these groups but does not provide complete coverage of the population at risk of COVID-19.  The coding is also incomplete with substantial missingness.  We have not merged ethnicity data from the Scottish Morbidity Record with name based classification as this would introduce differential misclassification of ethnicity by health status.  Access to primary care data would likely increase the completeness with which ethnicity is assigned but governance issues have not been resolved.

\newpage

# Supplementary tables
These tables including test-positive individuals who were not hospitalized are included by request.  They should be interpreted with caution, as selection for testing is influenced by many factors including health care worker status.

@@

pander(testpositives.ethnic.smr, missing=".",
       split.tables=Inf,
       caption="Severity of disease in test-positive cases, by ethnicity assigned in SMR records")

@@

```{r testpositives.carehome, echo=FALSE, warning=FALSE, message=FALSE}

pander(testpositives.carehome, missing=".",
       split.tables=Inf,
       caption="Care home status in test-positive cases, by ethnicity assigned from ONOMAP name classification")

```{r testpositives.healthboard, echo=FALSE, warning=FALSE, message=FALSE}

pander(testpositives.healthboard, missing=".",
       split.tables=Inf,
       caption="Health board of residence in test-positive cases, by ethnicity assigned from ONOMAP name classification")

## Relation of test-positive status to ethnicity in the Scottish population

```{r onomap.testpos, echo=FALSE, warning=FALSE, message=FALSE}

rownames(table.testpositives.demog) <- gsub("_", " ", rownames(table.testpositives.demog))
knitr::kable(table.testpositives.demog,
             escape=FALSE, 
             booktabs=TRUE,
             label="testpos.onomap",
             digits=c(0, 0, 0, 200, 0, 200), 
             col.names=c(clean.header(colnames(table.testpositives.demog)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of test-positive status with ethnicity based on name classification") %>%
    column_spec(column=1, width="4.5cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    landscape()
```

```{r ethnicsmr.testpos, echo=FALSE, warning=FALSE, message=FALSE}

rownames(table.testpositives.demog.ethnicsmr) <- gsub("_", " ", rownames(table.testpositives.demog.ethnicsmr))
knitr::kable(table.ethnicsmr.aug,
             escape=FALSE, 
             booktabs=TRUE,
             label="testpos.ethnicsmr",
             digits=c(0, 0, 0, 200, 0, 200), 
             col.names=c(clean.header(colnames(table.testpositives.demog.ethnicsmr)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of test-positive status with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
    column_spec(column=1, width="4.5cm")
```
